---
title: "Figures for quantile niche manuscript"
date: "`r Sys.Date()`"
author: "Sarah Goslee"
output:
  rmarkdown::pdf_document:
geometry: margin=1in
---


# Manuscript figures

This file produces the figures for the manuscript:

What Could Be: Quantile Random Forest Models of the Abundance-Suitability Relationship
Sarah C. Goslee1, Martina Calovi2, Melanie Kammerer1, Darin J. McNeil3, and Douglas A. Miller4
1. USDA-ARS Pasture Systems and Watershed Management Research Unit, University Park, PA 16802, USA
2. Department of Geography, Norwegian University of Science and Technology, Trondheim, Norway
3. Department of Environmental Sciences, University of North Carolina Wilmington, Wilmington, NC 28403, USA
4. Department of Ecosystem Science and Management, The Pennsylvania State University, University Park, PA 16802, USA


```{r setup}

    library(sp)
    library(raster)

    load("data/statesLL.RDA")
    load("data/PRISMgridLL.RDA")

    example.species <- c("Acer.rubrum", "Picea.rubens")
    example.colors  <- c("#1f78b4", "#33a02c")

    resolution <- 300


```


# Figure 1.

Species potential abundance distribution along an environmental gradient is not necessarily symmetrical, or even continuous. Observed abundances at any point on the gradient fall along a distribution from 0 to the potential abundance. If an observed abundance is greater than the expected potential abundance, it may be a sink site, or the species may be supported by unmodeled factors (for instance management).

```{r fig1}


    tiff("fig1.tif", width=4*resolution, height=3*resolution, pointsize=resolution/8, compression="lzw")

    par(mai = par()$mai * c(1, 1, .3, 1))

    #####

    # generate artificial abundance-suitability curve
    set.seed(1111)
    maxabun.generate <- hist(rbeta(30000,2,5), breaks=100, plot = FALSE)

    maxabun <- data.frame(
        x = c(0, 2, 2 + (maxabun.generate$breaks * 6)[-1], 7.76, 10),
        y = c(0, 0, smooth.spline(70 * maxabun.generate$counts / max(maxabun.generate$counts))$y, 0, 0))

    # generate 100 random points under the curve
    sample.points <- t(sapply(seq_len(100), function(i) {
        thisrow <- sample(seq_len(nrow(maxabun))[maxabun$y > 0.2], 1)
        c(maxabun$x[thisrow], maxabun$y[thisrow] * runif(1))
    }))
    sample.points <- data.frame(sample.points)
    colnames(sample.points) <- c("x", "y")

#####

    plot(c(0, 10), c(0, 100), xlab = "Environmental gradient", ylab = "Species abundance", main = "", type = "n")
    lines(maxabun$x, maxabun$y, lwd = 3, col = "#d95f02")

    points(sample.points)

    # add 2 sink points above the curve

    points(3.2, 80, pch = 8)
    text(3.2, 80+3, pos = 3, "Greater than\nexpected abundance")

    points(1, 15, pch = 8)
    text(1, 15+3, pos = 3, "Outside of\nexpected\nenvironment")

    # illustrate sampling data vs suitability

    slice.min <- 4.46
    slice.max <- 4.76

    segments(slice.min, 0, slice.min, maxabun$y[maxabun$x == slice.min])
    segments(slice.max, 0, slice.max, maxabun$y[maxabun$x == slice.max])

    sample.points.sub <- subset(sample.points, x >= slice.min & x <= slice.max)
    points(sample.points.sub, pch = 16)

    arrows(mean(slice.min, slice.max) + .3, with(maxabun, mean(y[x >= slice.min & x <= slice.max])), 6, with(maxabun, mean(y[x >= slice.min & x <= slice.max])), lwd = 2, col = "#d95f02", code = 1)
    text(5.8, with(maxabun, mean(y[x >= slice.min & x <= slice.max])) + 3, pos = 4, "Potential abundance", col = "#d95f02")

    arrows(7.6, min(sample.points.sub$y), 7.6, max(sample.points.sub$y), code = 3)
    text(7.6, 20, pos = 4, "Observed\nabundances")

    dev.off()

```



# Figure 2

Panels a and c show the observed presence distribution of seedlings of two tree species, one with a very broad and one with a narrower range, superimposed on gray shading of the FIA sampling extent. Panels b and d show the corresponding 95% quantile predictions; darker shades denote greater suitability.  


```{r fig2}

## distribution maps

plot.distr <- function(thissp, thiscol = "black", label) {

	plot(states)
	points(crd[, 1:2], col="gray85", pch = ".")
	points(fia$longitude[grepl(thissp, fia$species) & fia$scaledAbundance > 0], fia$latitude[grepl(thissp, fia$species) & fia$scaledAbundance > 0], pch = ".", col=thiscol)
	plot(states, add = TRUE)
    spname <- sub("\\.", " ", thissp)

    if(missing(label)) {
	    title(bquote(italic(.(spname)) ~ "distribution"), col.main = thiscol)
    } else {
	    title(bquote(.(label) ~ italic(.(spname)) ~ "distribution"), col.main = thiscol)
    }

	invisible()

}

## prediction maps in geographic space

plot.map <- function(thissp, thiscol = "black", type = "mean", label) {

	load(file.path("models", "spatial", paste("spatial", thissp, "RDA", sep=".")))
	pq.map@data[pq.map@data < 1] <- NA

	if(type == "mean") {
	    image(pq.map, "mean", col = rev(scico::scico(30, palette = "oslo")), main = "", zlim=c(0, max(c(pq.map$mean, pq.map$q95), na.rm = TRUE)), xlim = range(crd$longitude), ylim = range(crd$latitude))
		plot(states, add=TRUE)
        spname <- sub("\\.", " ", thissp)

        if(missing(label)) {
    	    title(bquote(italic(.(spname)) ~ "mean"), col.main = thiscol)
        } else {
    	    title(bquote(.(label) ~ italic(.(spname)) ~ "mean"), col.main = thiscol)
        }
	
    }
	if(type == "q95") {
	    image(pq.map, "q95", col = rev(scico::scico(30, palette = "davos")), main = "", zlim=c(0, max(c(pq.map$mean, pq.map$q95), na.rm = TRUE)), xlim = range(crd$longitude), ylim = range(crd$latitude))
		plot(states, add=TRUE)
        spname <- sub("\\.", " ", thissp)

        if(missing(label)) {
    	    title(bquote(italic(.(spname)) ~ "95%"), col.main = thiscol)
        } else {
    	    title(bquote(.(label) ~ italic(.(spname)) ~ "95%"), col.main = thiscol)
        }
	
    }

	invisible()

}

######

    tiff("fig2.tif", width=5*resolution, height=4*resolution, pointsize=resolution/8, compression="lzw")

    par(mfcol = c(2, 2))

    # distribution panels

    thislab <- c("a.", "c.")

    par(mai = c(0, 0, par()$mai[3], 0))

        for(i in 1:2) {
            thissp  <- example.species[i]
            thiscol <- example.colors[i]
            plot.distr(thissp, thiscol, label = thislab[i])
        }

    # quantile model panels

    thislab <- c("b.", "d.")

        for(i in 1:2) {
            thissp  <- example.species[i]
            thiscol <- example.colors[i]
            plot.map(thissp, thiscol, "q95", label = thislab[i])
        }

    dev.off()


```


# Figure 3




```{r fig3}


plot.pred <- function(thissp, thiscol = "black", legend = TRUE, label) {

	load(file.path("models", "envpred", paste("envpred", thissp, "RDA", sep=".")))

    plot(fia99.prism[[thissp]], datp.envpredq$predictions[, 7], col="#d95f02", pch = 2, xlab = "Measured abundance", ylab = "Modeled abundance", xlim = c(0, max(datp.envpredq$predictions[, 7])), ylim = c(0, max(datp.envpredq$predictions[, 7])))
    points(fia99.prism[[thissp]], datp.envpred$predictions, col="#7570b3", pch = 1)
	abline(c(0, 1), lwd = 9)

    spname <- sub("\\.", " ", thissp)

    if(missing(label)) {
	    title(bquote(italic(.(spname))), col.main = thiscol)
    } else {
    	title(bquote(.(label) ~ italic(.(spname))), col.main = thiscol)
    }
	
	if(legend) legend("topleft", legend = c("RF", "qRF, 95%"), pch = c(1, 2), col = c("#7570b3", "#d95f02"), cex=.9)

	invisible()

}


###

tiff("fig3.tif", width=2.5*resolution, height=5.5*resolution, pointsize=resolution/7, compression="lzw")

### predictions

par(mfcol=c(2, 1))
par(mai = par()$mai * c(1, 1, .9, 1))

thislab <- c("a.", "b.")

	for(i in 1:2) {
		thissp  <- example.species[i]
		thiscol <- example.colors[i]
		plot.pred(thissp, thiscol, legend = c(FALSE, TRUE)[i], label = thislab[i])
	}

dev.off()


```


# Figure 4


```{r fig4}


# smoothed line plots for environmental gradient

qline <- function(x, y, n = 50) {
	breaks <- seq(min(x), max(x), length.out = n)
	groups <- cut(x, breaks = breaks)
	sx <- sy <- rep(NA, length(levels(groups)))
	for(i in seq_along(levels(groups))) {
		sx[i] <- mean(breaks[c(i, i+1)])
		tempy <- y[groups == levels(groups)[i]]
		tempy <- tempy[!is.na(tempy) & tempy > 0]
		sy[i] <- quantile(tempy, .99)
	}
	sy[is.na(sy)] <- 0
	list(x = sx, y = sy, sy = smooth(sy))
}


# full set of quantile lines

plot.qlines <- function(thissp, thiscol = "black", legend = TRUE, label) {

	load(file.path("models", "envpred", paste("envpred", thissp, "RDA", sep=".")))
	load(file.path("models", "mqpred", paste("mqpred", thissp, "RDA", sep=".")))


    # aggregated fia samples
	spdat <- data.frame(datp.envpred$predictions, datp.envpredq$predictions, datp.envpredqs$predictions)
	colnames(spdat) <- c("mean", "q05", "q10", "q25", "q50", "q75", "q90", "q95", "s05", "s10", "s25", "s50", "s75", "s90", "s95")
	spdat[spdat < 0] <- 0
	spdat <- data.frame(prismid = env.prism$prismid, y = fia99.prism[[thissp]], spdat)
    spdat <- data.frame(rowid = seq_len(nrow(spdat)), spdat)
    spdat <- merge(spdat, env.gradient, all.x = TRUE)
    spdat <- spdat[order(spdat$rowid), ]
	spdat <- spdat[spdat$y > 0, ]


	mqdat <- data.frame(datp.pred, datp.predq, datp.predqs)
	colnames(mqdat) <- c("mean", "q05", "q10", "q25", "q50", "q75", "q90", "q95", "s05", "s10", "s25", "s50", "s75", "s90", "s95")
	mqdat[mqdat < 0] <- 0
	mqdat <- data.frame(mqdat, env.gradient)

	#

	plot(mqdat$Comp.1, mqdat$q95, type = "n", xlab = "Environmental gradient: PCA 1", ylab = "Abundance")
    spname <- sub("\\.", " ", thissp)

    if(missing(label)) {
	    title(bquote(italic(.(spname))), col.main = thiscol)
    } else {
    	title(bquote(.(label) ~ italic(.(spname))), col.main = thiscol)
    }

	thisline <- qline(x = mqdat$Comp.1, y = mqdat$q05, n = 40)
	with(thisline, lines(x, sy, col = "gray70", lwd = 5))

	thisline <- qline(x = mqdat$Comp.1, y = mqdat$q10, n = 40)
	with(thisline, lines(x, sy, col = "gray60", lwd = 5))

	thisline <- qline(x = mqdat$Comp.1, y = mqdat$q25, n = 40)
	with(thisline, lines(x, sy, col = "gray50", lwd = 5))

	thisline <- qline(x = mqdat$Comp.1, y = mqdat$q50, n = 40)
	with(thisline, lines(x, sy, col = "gray40", lwd = 5))

	thisline <- qline(x = mqdat$Comp.1, y = mqdat$q75, n = 40)
	with(thisline, lines(x, sy, col = "gray30", lwd = 5))

	thisline <- qline(x = mqdat$Comp.1, y = mqdat$q90, n = 40)
	with(thisline, lines(x, sy, col = "gray20", lwd = 5))

	thisline <- qline(x = mqdat$Comp.1, y = mqdat$q95, n = 40)
	with(thisline, lines(x, sy, col = "gray10", lwd = 5))

	thisline <- qline(x = mqdat$Comp.1, y = mqdat$mean, n = 40)
	with(thisline, lines(x, sy, col = "#7570b3", lwd = 13))

	if(legend) legend("topright", col=rev(c("gray70", "gray60", "gray50", "gray40", "gray30", "gray20", "gray10", "#7570b3")), lty=1, legend = rev(c("5%", "10%", "25%", "50%", "75%", "90%", "95%", "mean")), cex=.95, lwd=rev(c(rep(5, 7), 9)))

	invisible()

}


####

tiff("fig4.tif", width=2.5*resolution, height=5.5*resolution, pointsize=resolution/8, compression="lzw")

### quantile lines

par(mfcol=c(2, 1))
par(mai = par()$mai * c(1, 1, .9, 1))
thislab <- c("a.", "b.")

	for(i in 1:2) {
		thissp  <- example.species[i]
		thiscol <- example.colors[i]
		plot.qlines(thissp, thiscol, legend = c(FALSE, TRUE)[i], label = thislab[i])
	}

dev.off()



```


# Figure 5



```{r fig5}


trimmedabline <-  function(thislm, ...)  {

	ends <- predict(thislm, newdata = data.frame(x  =  c(0, 10000)))
	segments(0, ends[1], 10000, ends[2], ...)

	invisible()

}

####

tiff("fig5.tif", width=4*resolution, height=4*resolution, pointsize=resolution/7, compression="lzw")
par(mai = par()$mai * c(1, 1, .3, 1))

plot(c(0, 7000), c(0, 7000), type="n", xlab="Measured abundance",  ylab="Modeled abundance", main = "")
legend("topleft", legend = c("RF", "95% quantile RF"), col=c("#7570b3", "#d95f02"), lty = c(1, 3), lwd=12)

for(i in seq_along(splist)) {
	
	thissp <- splist[i]

	plotdat <- data.frame(
		x    = fia99.prism[[thissp]], 
		y    = all.pred[[thissp]],
		y95  = all.pred95[[thissp]])

	trimmedabline(lm(y    ~ x, data = plotdat), col = "#7570b3", lty = 1, lwd = 3)
	trimmedabline(lm(y95  ~ x, data = plotdat), col = "#d95f02", lty = 3, lwd = 3)

}
abline(c(0, 1), lwd = 9)

dev.off()


```

# Supplementary figure 1


Aggregation on maximum instead of mean.


```{r suppl1}



### Figure: quantile on mean vs quantile on max RF prediction lines

tiff("figS1.tif", width=6*resolution, height=6*resolution, pointsize=.025*6*resolution)

plot(c(0, 7000), c(0, 7000), type="n", xlab="Measured abundance",  ylab="Modeled abundance", main = "")
legend("bottomright", legend = c("95% quantile, data aggregated by mean", "95% quantile, data aggregated by max"), col=c("#d95f02", "darkgray"), lty = c(1, 3), lwd = 12)

for(i in seq_along(splist)) {
	
	thissp <- splist[i]

	plotdat <- data.frame(
		x    = fia99.prism[[thissp]], 
		y    = all.pred[[thissp]],
		y95  = all.pred95[[thissp]],
		ym   = all.predm[[thissp]],
		ym95 = all.predm95[[thissp]])


	trimmedabline(lm(y95  ~ x, data = plotdat), col = "#d95f02", lty = 1, lwd = 3)
	trimmedabline(lm(ym95 ~ x, data = plotdat), col = "darkgray", lty = 3, lwd = 3)

}
abline(c(0, 1), lwd = 9)

dev.off()

```


# Figure S2

Smoothed quantiles


```{r figS2}


thissp <- "Acer.rubrum"
thisfiles <- load(paste("models/envpred/envpred", thissp, "RDA", sep="."))
datp.envpredqs$predictions[datp.envpredqs$predictions < 0] <- 0

tiff("figS2.tif", width=7*resolution, height=7*resolution, pointsize=resolution/6)

### quantile lines

par(mfcol=c(2, 2))

plot(fia99.prism[[thissp]], datp.envpredq$predictions[, 2], col = "#d95f02", cex=.5, xlab = "Measured abundance", ylab = "10% quantile RF", main = paste(thissp, "quantile RF"))
plot(fia99.prism[[thissp]], datp.envpredqs$predictions[, 2], col = "#d95f02", cex=.5, xlab = "Measured abundance", ylab = "10% quantile RF", main = paste(thissp, "smoothed quantile RF"))


plot(fia99.prism[[thissp]], datp.envpredq$predictions[, 7], col = "#d95f02", cex=.5, xlab = "Measured abundance", ylab = "95% quantile RF", main = paste(thissp, "quantile RF"))
plot(fia99.prism[[thissp]], datp.envpredqs$predictions[, 7], col = "#d95f02", cex=.5, xlab = "Measured abundance", ylab = "95% quantile RF", main = paste(thissp, "smoothed quantile RF"))

dev.off()


rm(list = thisfiles)
rm(thisfiles, thissp)


```



